\section{Conclusion}
\label{sec:conclusion}

We presented SA-NAS, a Structure-Aware Neural Architecture Search
framework for Mixture-of-Experts models, which couples per-expert
architecture search with the cluster structure of the underlying
data. The framework relies on a Graph Neural Network surrogate that
predicts validation accuracy from an architecture description and a
binary cluster mask, trained inside an active-learning loop with
Monte-Carlo dropout uncertainty estimates. Three optimisation
algorithms (sampling, Gumbel-Softmax gradient and EM) instantiate the
same cluster-aware MoE objective and share a common implementation
of the surrogate.

On the controlled toy problem the EM variant recovers the ideal
linear/ring split with $K = 2$ experts, confirming that the
formulation is well posed when the surrogate is reliable. On a
deliberately heterogeneous CIFAR-100 / SVHN mixture, whose
ground-truth domains give a sharp oracle partition, SGEM recovers that
partition on $92\%$ of clusters \emph{without ever seeing the source
labels}: one expert specialises to a pure CIFAR sub-task while the
other collects the SVHN clusters. Under the resulting hard cluster gate
it attains $0.620$ test accuracy, ahead of the random-MoE baseline
($0.547$) and of $K$ copies of a single DARTS architecture ($0.591$),
and within $1.3$~p.p.\ of an oracle MoE ($0.633$) that is handed the
true split. A diagnostic analysis was decisive: an earlier pipeline
that scored experts on the whole validation set and sampled cluster
masks uniformly never exposed the surrogate to near-pure subsets and
collapsed the routing, whereas region-restricted evaluation together
with a domain-agnostic log-uniform mask sampler removes this bias and
lets the cluster-aware objective discover the data structure on its
own.

These results are obtained on a two-domain benchmark with $K = 2$ and
a modest number of seeds; broadening to more domains, larger $K$, and a
wider seed sweep is the natural next step. We also plan to scale the
search space, to extend the framework to transformer-based MoE layers,
and to strengthen the theory for the cluster-aware objective --- in
particular by folding the load-balancing term, which the present
convergence analysis omits, into the guarantee.
