\section{Related Work}
\label{sec:related_work}

\noindent

\vspace{2mm}

\textbf{Mixture-of-Experts.}
The Mixture-of-Experts (MoE) framework was originally introduced
in~\cite{jacobs1991adaptive,jordan1994hierarchical,jacobs1997bayesian}
as a way to combine specialised statistical models through a learnable
gating mechanism. Subsequent theoretical analyses characterised
identifiability and convergence in classical MoE
formulations~\cite{nguyen2023general,nguyen2023demystifying,piwko2025divide,rasmussen2001infinite}.
With the rise of deep learning, MoE has been revived in the form of
sparsely activated neural networks, where the gate routes each input
to only a small subset of
experts~\cite{eigen2013learning,shazeer2017outrageously,riquelme2021scaling,fedus2022switch}.
This sparse formulation enables training of models with hundreds of
billions of parameters while keeping per-token compute roughly
constant, and is now a standard building block of modern large-scale
transformers~\cite{jiang2024mixtral,dai2024deepseekmoe,qiu2025demons}.

\vspace{2mm}

\textbf{Neural Architecture Search.}
Neural Architecture Search (NAS) automates the design of neural
networks and has matured into a wide research
field~\cite{ren2021comprehensive}. Many-shot approaches based on
reinforcement learning~\cite{zoph2016neural,baker2016designing,zoph2018learning}
or evolutionary search~\cite{real2019regularized,yang2020cars}
explicitly train candidate architectures and select the best
performer. One-shot methods reduce the search cost by training a
single supernet that contains all candidate architectures as
sub-networks~\cite{liu2018darts,chen2020stabilizing,you2020greedynas,nayman2019xnas}.
DARTS~\cite{liu2018darts} is the most prominent representative of this
family and forms one of the baselines we compare against in our
experimental study.

\vspace{2mm}

\textbf{Neural Architecture Search for Mixture-of-Experts.}
The intersection of NAS and MoE is a comparatively recent and still
sparsely populated research area. AutoMoE~\cite{jawahar2022automoe}
applies an evolutionary search on top of a shared supernet to obtain
heterogeneous MoE architectures for neural machine translation.
CMN~\cite{han2024cmn} co-designs MoE topologies for
computing-in-memory hardware. The work most closely related to ours
is MoE-NAS~\cite{mecharbat2025moenas}, which optimises the number of
experts per layer but keeps the expert architectures shared. In
contrast, we explicitly search over per-expert architectures and tie
this search to the cluster structure of the input data.

\vspace{2mm}

\textbf{Theoretical Understanding.}
A growing body of theoretical work studies why MoE models perform
well on data with strong cluster or modality structure.
\cite{chen2022towards} proves on simplified settings that MoE
provably benefits from clusterable data, while~\cite{li2024theory}
analyses the optimisation landscape of sparse gates. Empirical
evidence from multimodal and multi-task
learning~\cite{chai2022mixture,guo2018multi,hu2025multi,zhou2022mixture}
further supports the claim that experts naturally specialise to
distinct sub-distributions. Our work builds on this perspective and
turns it into an explicit search objective by optimising over
\emph{per-cluster} expert architectures rather than treating the
cluster structure as a by-product of training.

\vspace{2mm}
